---
title: "Extracting and Viewing Extra V and J Segments"
author: "Wes Horton"
date: "May 5, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem Summary

In our MiXCR export clones files, there are a handful of V and J regions that are called that we don't have primers for. This means that either MiXCR is mistakenly mapping one of our 20 V or 13 J regions to a different region, or some of our primers are amplifying extra V and J regions.  

Is this a problem that we need to address, or is the impact small enough that we can confidently ignore it without jeopardizing our results?

```{r, echo=FALSE}
# Set spike dir
spike.dir <- "~/Desktop/OHSU/tcr_spike/"
# Set clone dir
clone.dir <- "~/Desktop/OHSU/tcr_spike/data/DNA151124LC/raw_clones/"

# Function to clean mixcr data frame
clean <- function(mixcr.df){
  # Remove extraneous characters
  mixcr.df$"V.segments" <- sub("TRB", "", mixcr.df$"Best.V.hit");
  mixcr.df$"J.segments" <- sub("TRB", "", mixcr.df$"Best.J.hit");
  mixcr.df$"V.segments" <- sub("\\*00", "", mixcr.df$"V.segments");
  mixcr.df$"J.segments" <- sub("\\*00", "", mixcr.df$"J.segments");
  mixcr.df$"V.segments" <- gsub("-", "", mixcr.df$"V.segments")
  return(mixcr.df)
} # clean(mixcr.df)
```

```{r cars}
###
###  Read in spike identity table and extract V and J segment IDs
###  Read in clonotype files, create empty data frame to populate
###  Extract extra V and Js from clonotype files and populate data frame
###
# Read in spike identity table
spike.table <- read.delim(paste(spike.dir, "text_barcodesvj.txt", sep = ''),
                        header = T, sep = ' ')

# Remove trailing dashes for hygiene's sake. Also change V1212 (for consistency)
spike.table$V <- gsub("-","", spike.table$V)
spike.table$V <- gsub("V1212", "V121", spike.table$V)

# Extract the V and J sequences from spike table
vsegs <- unique(spike.table$V)
jsegs <- unique(spike.table$J)
  
# Read in normalized clone files and order them
all.clones <- list.files(clone.dir)
all.clones <- all.clones[order(as.numeric(gsub(".*_S|_alignment_.*", '', all.clones)))]

# Create empty data frame to populate
extra.segs.df <- data.frame()
summary.extra.segs.df <- data.frame(matrix(nrow = 0, ncol = 7))


# Populate summary data frame
for (i in 1:length(all.clones)){
  # Read in clone file
  curr.df <- read.table(file.path(clone.dir, all.clones[i]), sep = '\t', header = T)
  # Clean up segments
  curr.df <- clean(curr.df)
  
  # Extract V and J segments not in our 260
  curr.extra.V <- curr.df[!(curr.df$V.segments %in% vsegs),]
  curr.extra.J <- curr.df[!(curr.df$J.segments %in% jsegs),]
  curr.extras <- rbind(curr.extra.V, curr.extra.J)

  # Create vector of sample name to combine with df
  sample <- rep(strsplit(all.clones[i], split = "_")[[1]][3], length(curr.extras[,1]))
  
  # Combine
  curr.extras <- cbind(sample, curr.extras)
  
  # Add to overall df
  extra.segs.df <- rbind(extra.segs.df, curr.extras)
  
  # Populate Summary Table
  num.clone <- as.numeric(length(curr.extras$sample))
  percent.clone <- as.numeric(num.clone / length(curr.df$Clone.count) * 100)
  num.count <- as.numeric(sum(curr.extras$Clone.count))
  percent.count <- as.numeric(num.count / sum(curr.df$Clone.count) * 100)
  vs <- as.character(unique(curr.extra.V$V.segments))
  vs <- paste(vs, collapse = ' ')
  js <- as.character(unique(curr.extra.J$J.segments))
  js <- paste(js, collapse = ' ')
  
  new.row <- c(sample[1], num.clone, percent.clone, num.count, percent.count, vs, js)
  
  summary.extra.segs.df <- rbind(summary.extra.segs.df, new.row, stringsAsFactors = F)
  
}  #  for

colnames(summary.extra.segs.df) <- c("Sample", "Number of clonotypes", "Percent of Total Clonotypes", 
                                     "Clonotype Count", "Percent of Total Count", "Extra Vs",
                                     "Extra Js")
```

## Evaluation

Let's look at a brief summary of the percent columns to see how big of an impact this is

```{r }
summary(as.numeric(summary.extra.segs.df$`Percent of Total Clonotypes`))
summary(as.numeric(summary.extra.segs.df$`Percent of Total Count`))

```


