---
title: "MiXCR_batch_compare"
author: "Wes Horton"
date: "June 1, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(grid)
source("http://peterhaschke.com/Code/multiplot.R")
```

## Objective

We're having issues with out MiXCR output. Different runs of the 160107 batch were compared in the MiXCR\_QC\_analysis.Rmd, but this will compare the different batches, all run using the same parameters.

```{r, echo = FALSE}
# Read in align and assemble QC files

### Align ###

# 150624
align.150624 <- read.delim("~/Desktop/OHSU/tcr_spike/data/equiv_DNA150624LC/QC/mixcr.alignment.QC.summary.txt", sep = ',', header = T)
sample <- as.numeric(gsub(".*S|.assembled.*", '', align.150624$inputs))
batch <- rep("150624", length(align.150624$analysis.date))
align.150624 <- cbind(sample, batch, align.150624)

# 150826
align.150826 <- read.delim("~/Desktop/OHSU/tcr_spike/data/equiv_DNA150826LC/QC/mixcr.alignment.QC.summary.txt", sep = ',', header = T)
sample <- as.numeric(gsub(".*S|.assembled.*", '', align.150826$inputs))
batch <- rep("150826", length(align.150826$analysis.date))
align.150826 <- cbind(sample, batch, align.150826)
align.150826 <- align.150826[!(align.150826$sample == "141"),]

# 151124
align.151124 <- read.delim("~/Desktop/OHSU/tcr_spike/data/equiv_DNA151124LC/QC/mixcr.alignment.QC.summary.txt", sep = ',', header = T)
sample <- as.numeric(gsub(".*S|.assembled.*", '', align.151124$inputs))
batch <- rep("151124", length(align.151124$analysis.date))
align.151124 <- cbind(sample, batch, align.151124)
align.151124 <- align.151124[!(align.151124$sample == "142"),]

# 160107
align.160107 <- read.delim("~/Desktop/OHSU/tcr_spike/data/equiv_DNA160107LC/QC/mixcr.alignment.QC.summary.txt", sep = ',', header = T)
sample <- as.numeric(gsub(".*S|.assembled.*", '', align.160107$inputs))
batch <- rep("160107", length(align.160107$analysis.date))
align.160107 <- cbind(sample, batch, align.160107)
align.160107 <- align.160107[!(align.160107$sample == "142"),]

# Combine together
align <- rbind(align.150624, align.150826, align.151124, align.160107)

### Assemble ###
# 150624
assemble.150624 <- read.delim("~/Desktop/OHSU/tcr_spike/data/equiv_DNA150624LC/QC/mixcr.assemble.QC.summary.txt", sep = ',', header = T)
sample <- as.numeric(gsub(".*S|_align.*", '', assemble.150624$inputs))
batch <- rep("150624", length(assemble.150624$analysis.date))
assemble.150624 <- cbind(sample, batch, assemble.150624)

# 150826
assemble.150826 <- read.delim("~/Desktop/OHSU/tcr_spike/data/equiv_DNA150826LC/QC/mixcr.assemble.QC.summary.txt", sep = ',', header = T)
sample <- as.numeric(gsub(".*S|_align.*", '', assemble.150826$inputs))
batch <- rep("150826", length(assemble.150826$analysis.date))
assemble.150826 <- cbind(sample, batch, assemble.150826)
assemble.150826 <- assemble.150826[!(assemble.150826$sample == "141"),]

# 151124
assemble.151124 <- read.delim("~/Desktop/OHSU/tcr_spike/data/equiv_DNA151124LC/QC/mixcr.assemble.QC.summary.txt", sep = ',', header = T)
sample <- as.numeric(gsub(".*S|_align.*", '', assemble.151124$inputs))
batch <- rep("151124", length(assemble.151124$analysis.date))
assemble.151124 <- cbind(sample, batch, assemble.151124)
assemble.151124 <- assemble.151124[!(assemble.151124$sample == "142"),]

# 160107
assemble.160107 <- read.delim("~/Desktop/OHSU/tcr_spike/data/equiv_DNA160107LC/QC/mixcr.assemble.QC.summary.txt", sep = ',', header = T)
sample <- as.numeric(gsub(".*S|_align.*", '', assemble.160107$inputs))
batch <- rep("160107", length(assemble.160107$analysis.date))
assemble.160107 <- cbind(sample, batch, assemble.160107)
assemble.160107 <- assemble.160107[!(assemble.160107$sample == "142"),]

# Combine together
assemble <- rbind(assemble.150624, assemble.150826, assemble.151124, assemble.160107)

# Calculate percent of clonotypes eliminated by PCR Error Correction
assemble$total.pot.clones <- assemble$clonotype.count + assemble$clonotypes.elim.error.corr
assemble$pct.lost.to.pcr.corr <- assemble$clonotypes.elim.error.corr / assemble$total.pot.clones * 100
```

### Results

We've generally had pretty good alignment (based on observations from the 160107 batch), with most failed alignments due to bad J reads. Let's see how this looks for all of the batches:

```{r, echo = FALSE}
pct.align <- ggplot(align, aes(x = batch, y = aligned.pct, group = batch)) +
  geom_boxplot(aes(fill = batch)) +
  ylab("Percent Aligned of Total") +
  xlab("Batch ID") +
  ggtitle("Fig. 1 Successfully Aligned Reads")

pct.align
```

Alignment percentages are relatively consistent. Turns out the 160107 batch was the worst of them. Let's look at the reasons for failure:

```{r, echo = FALSE}
pct.low.score <- ggplot(align, aes(x = batch, y = failed.alignment.low.score,
                                   group = batch)) +
  geom_boxplot(aes(fill = batch)) +
  ylab("Percent Not Aligned of Total") +
  xlab("Batch ID") +
  ggtitle("Fig. 2 Failed Alignments Due to Low Score")

failed.j <- ggplot(align, aes(x = batch, y = failed.alignment.j.hits,
                              group = batch)) +
  geom_boxplot(aes(fill = batch)) +
  ylab("Percent Not Aligned of Total") +
  xlab("Batch ID") +
  ggtitle("Fig. 3 Failed Alignments Due to Absence of J Hit")

pct.low.score

failed.j
```

Everything looks pretty standard between the different batches. Let's look at the assembly statistics:

```{r, echo = F}
pct.assemble <- ggplot(assemble, aes(x = batch, y = pct.total.reads.used, 
                                     group = batch)) +
  geom_boxplot(aes(fill = batch)) +
  ylab("Percent Reads Used of Total") +
  xlab("Batch ID") +
  ggtitle("Fig. 4 Percent Assembled Reads")

pct.core <- ggplot(assemble, aes(x = batch, y = pct.reads.used.as.core, group = batch)) +
  geom_boxplot(aes(fill = batch)) +
  ylab("Reads Used for Core (Percent of Reads Used)") + 
  xlab("Batch ID") +
  ggtitle("Fig. 5 Assembled Reads Used as Core Clones")

pct.mapped <- ggplot(assemble, aes(x = batch, y = pct.reads.used.mapped.low.quality,
                                   group = batch)) +
  geom_boxplot(aes(fill = batch)) +
  ylab("Reads Successfully Mapped (Percent of Used)") +
  xlab("Batch ID") +
  ggtitle("Fig. 6 Low Quality Reads Mapped to Core Clones")

pct.failed.mapped <- ggplot(assemble, aes(x = batch, y = pct.reads.dropped.failed.mapping,
                                          group = batch)) +
  geom_boxplot(aes(fill = batch)) +
  ylab("Reads Not Mapped (Percent of Used)") +
  xlab("Batch ID") +
  ggtitle("Fig. 7 Low Quality Reads Failing to Map")

pct.no.clonotype <- ggplot(assemble, aes(x = batch, 
                                         y = pct.reads.dropped.no.clonal.sequence,
                                         group = batch)) +
  geom_boxplot(aes(fill = batch)) +
  ylab("Percent Dropped of Total") + 
  xlab("Batch ID") +
  ggtitle("Fig. 8 Reads Dropped Due to No Clonal Sequence")

clone.elim.error <- ggplot(assemble, aes(x = pct.lost.to.pcr.corr)) +
  geom_histogram(aes(fill = batch)) +
  facet_grid(batch ~ .) +
  xlab("Percent lost to PCR Error Correction") + 
  ggtitle("Fig. 9 Clonotypes eliminated by PCR Error Correction")

pct.assemble

pct.core

pct.mapped

pct.failed.mapped

pct.no.clonotype

clone.elim.error
```