
title: "Independence Test for F and R Primers cnt'd - clustered heatmaps""
author: "Burcu"
date: "June 24, 2016"
output: html_document

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reshape2)
library(MASS)
library(gplots)
library(ggplot2)
library(ggthemes)
library(viridis)
library(gganimate)
```


## Aggregate all samples into a single table

```{r cars}

# Set directory
count.dir <- "C:/Users/burcu/Desktop/TCR/knitr-git/norm_knitr/independence/160609_spike_counts_25bp/160609_only_spikes"

# Read in files and sort by sample number
all.counts <- list.files(count.dir)
all.counts <- all.counts[order(as.numeric(gsub(".*_S|.assembled.*", '', all.counts)))]

# Read in first file to start aggregate data frame
count.df <- read.table(file.path(count.dir, all.counts[1]), sep =',', header = T)
count.df <- count.df[,3:5]

# Combine spike counts for all files into 1 data frame
  # Columns are samples
  # Rows are spikes
for (i in 2:length(all.counts)){
  curr.df <- read.table(file.path(count.dir, all.counts[i]), sep = ',', header = T)
  count.df <- cbind(count.df, curr.df$spike.count)
}   #   for i in 2:length(new.counts)
colnames(count.df) <- c("V", "J", seq(1:length(all.counts)))
```

## chi square test and residual table

``` {r}

## Preperaing the chi square matrix (V as columns, J as rows)

vbyj.first<- acast(count.df, V~J, value.var = 14)

## Chi squared test

chi.first <- chisq.test(vbyj.first, correct = F)

## Creating residuals tables

prtype <- chi.first$residuals


dd.col <- as.dendrogram(hclust(dist(prtype)))
col.ord <- order.dendrogram(dd.col)

dd.row <- as.dendrogram(hclust(dist(t(prtype))))
row.ord <- order.dendrogram(dd.row)

```

## Creating animation of CLUSTERED heatmaps of pearson residual matrices of 20 samples to visualize the patterns


``` {r}

# Creating a facet of heat maps with all samples next to each other by using ggplot


pear.res<-function(colName)
{
  temp.chi <- chisq.test(acast(count.df, V~J,value.var = colName) , correct = F)
  temp.res<-temp.chi$residuals
#ORDERING STARTS HERE
    #temp.res.ord<-scale(temp.res)[col.ord, row.ord]
  temp.res.ord<-temp.res[col.ord, row.ord]
  c_names <- attr(temp.res.ord, "dimnames")
  
  temp.res.ord <- as.data.frame(temp.res.ord)
  temp.res.ord$V <- c_names[[1]]
  temp.res.ord$V<-with(temp.res.ord, factor(V, levels=V, ordered = TRUE))
#Should be ordered nicely, now melt it!  
  ret.res<- melt(temp.res.ord,id_vars=c("V"))
  ret.res$sample<-rep(colName,nrow(temp.res))
  ret.res
}


res<-lapply(colnames(count.df[ ,3:22]), pear.res)
res.bound<-do.call(rbind,res)
#res.bound<-merge(tissue.lookup, res.bound, by.x = "num", by.y="sample")
```

``` {r fig.show="animate"}

gg<-ggplot(res.bound, aes(x=variable, y=V, fill=value, frame=sample))
gg<-gg+geom_tile(color="white", size=0.1)
gg<-gg+scale_fill_gradient2()
gg<-gg+theme(axis.text=element_text(size=6))
gg<-gg+theme(axis.text.x = element_text(angle = 90))
#gg<-gg+facet_wrap(~sample, ncol=5)
gg_animate(gg)
gg

```
## Comments:

Here we see clustering created V-J blocks. The first 3 J primers, then the next 6, then the last 4, each behaves a bit like a block. Similarly the first 6 V primers, then the next 7, then the last 7,
also behave a bit like a block. Animation shows similarly behaving blocks across samples. For clustering "eucliedan distance" is used.

